% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
  stu]{apa7}
%
\PassOptionsToPackage{style=apa}{biblatex}
%\PassOptionsToPackage{indent=7ex}{parskip}
%\usepackage{indentfirst}
%\usepackage[margin=1in]{geometry}
%% right side page number
%\usepackage{fancyhdr}
%\fancypagestyle{plain}{
%\fancyhf{}
%\renewcommand{\headrulewidth}{0pt}
%\fancyhead[R]{\thepage}
%}
%\pagestyle{fancy}
%\fancyhf{}
%\renewcommand{\headrulewidth}{0pt}
%\fancyhead[R]{\thepage}
%% double spacing
%\usepackage{setspace}
%\doublespacing
\affiliation{Brian Lamb School of Communication}
\course{COM 212: Interpersonal communication}
\professor{Dr. Hayden Barber}
\duedate{March 31, 2022}
%
\title{Chronemics in CMC}
\author{Sean F. O'Donovan}
\date{}

\usepackage{amsmath,amssymb}
\usepackage{lmodern}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\hypersetup{
  pdftitle={Chronemics in CMC},
  pdfauthor={Sean F. O'Donovan},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}
\urlstyle{same} % disable monospaced font for URLs
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage[]{biblatex}
\addbibresource{comPaper1References.bib}

\begin{document}
\maketitle

\hypertarget{introduction}{%
\section{Introduction}\label{introduction}}

When I communicate with others using computer mediated communication
(instant messaging, sms, social networks, also known as cmc), I
frequently notice that the conversation has a rhythm and pace. For
example, responding quickly to messages says something different from
responding slowly. Another example is that texting someone at 6 in the
morning feels very different from texting them at midnight. In
interpersonal communication, this is called chronemics. The phenomenon I
am choosing to investigate in this paper is chronemics in computer
mediated communication.

To define this cmc chronemics clearly, chronemics must be defined. A
study by \textcite{walther95} is a great place to start here. They quote
Burgoon and Saine's 1978 textbook to define chronemics as ``how we
perceive, structure, and react to time and \ldots{} the messages we
interpret from such usage'' \autocite[qtd. as cited in][
356]{walther95}. They go on to explain that ``Temporal cues may be
perceived by users, when they detect---with some systems, explicitly
directed to detect---the arrival of a message, or the amount of time
between one message and the next'' \autocite[ 356]{walther95}. In this
case, we're using the same definition. To summarize, chronemics in this
paper means the frequency of communication, the pacing of the
conversation itself (which is sometimes the same as the frequency of
communication), and the time at which it takes place. This paper
concerns cmc chronemics, so these attributes will be applied to computer
mediated communication. My personal belief as I explore this topic is
that chronemics communicate a lot in computer mediated communication,
and to some extent take the place of nonverbal cues. Additionally, I
think chronemics can be used to pull empirical data about relationships
out of individual and aggregated conversations. This paper will begin by
examining the way researchers have studied the existence of chronemics
in cmc, and then discuss the ways researchers have been able to put
chronemics to use. Researchers began to document the existence of
chronemics in cmc in the 90s, and are still catching up to the pace of
communication today.

\hypertarget{useful-theories}{%
\section{Useful Theories}\label{useful-theories}}

To start examining the literature on computer-mediated chronemics, we
need to define and explain some general theories. The first and
traditional theory is cues-filtered-out theory. This theory was
introduced by \textcite{sproull91}, and is an extremely naive approach.
It argues that cmc and cyberspace exist without any non-verbal cues, and
therefore is a fundamentally different space. The main conclusion is
that there will be fewer and less intense relationships in cyberspace. A
side effect, according to the theory, is that relationships are less
hierarchical and more democratic; everyone can have an equal share of
the conversation. A good summary of the theory is that it views cmc as
``inherently impersonal'' \autocite[ 16]{kim05}. The cues-filtered-out
approach is the traditional approach that most other theories exist to
disprove, and most people who have experienced cyberspace in any
meaningful way would agree that this theory is seriously lacking.

The second theory to discuss is the social information processing (SIP)
perspective. It claims that people are able to edit how they present
themselves more, while people also give each other the benefit of the
doubt more in cyberspace \autocite{kim05}. This means that there is an
opportunity for even more emotion and intensity of relationship than
normal in cyberspace. It also claims that because people are missing
normal cues, they fixate on what cues there are, such as writing style,
written ``nonverbals'' like emojis, and other text cues like repeated
letters \autocite{kim05}. \textcite{walther96} uses longitudinal studies
to explain both why this SIP perspective is the right one, and why
studies supporting cues-filtered-out theory delivered the results they
did. In short, cmc can be slower both to communicate in at all, and
therefore to develop relationships with \autocite[ 12]{walther96}. When
cues-filtered-out experiments convened ``one-shot'' partnerships or
groups, this failed to account for the time which relationships need to
develop \autocite[ 12]{walther96}.

\hypertarget{describing-cmc-chronemics-phenomena}{%
\section{Describing CMC Chronemics
Phenomena}\label{describing-cmc-chronemics-phenomena}}

The first and most important discovery by researchers about cmc
chronemics was the acknowledgement that they exist and influence
people's impression of the conversation. This is the claim made by
\textcite{walther95}, in contrast to the previous view that cmc is
impersonal and doesn't contain nonverbal cues. This was the first, most
important paper which refuted cues-filtered-out theory. It finds that
altering time and date stamps changes the way external observers see the
communication. \textcite{walther95} validates two hypotheses and has
complex results for others. First, it proves that a social message sent
at night signals less dominance than a social message sent during the
day and that a task message sent at night is more dominant than a task
message sent during the day. This suggests that even in cmc, ``different
parts of the day correspond to different activity contexts, and formal
`business hours' versus informal `after hours' carry different
expectations'' \autocite[ 361]{walther95}. Second, it proves that a slow
reply to a social message indicates greater affection than a fast reply,
but a slow reply to a task message indicates less affection than a fast
reply \autocite[ 370]{walther95}. Not all of their hypotheses have
simple results.

One complex finding is that a faster response suggests a more equal
footing between participants, rather than suggesting that the replier is
less dominant than the sender \autocite[ 371]{walther95}. Something
similar is pointed out in \textcite{kalman13}, which will be discussed
as well. \textcite{walther95} was originally based on email
communication, but it was also effectively replicated for sms by
\textcite{doring09}.

The next important phenomenon documented by researchers is that of
online silence. The idea that even in asynchronous communications, it is
possible for silence to exist was documented by \textcite{ravid} and
also by \textcite{kalman05}. Kalman explains that ``silence can be
defined as no response after an x period of time, at which, say, 99\% or
97\% of the responses have already been created'' (2005).
\textcite{kalman05} also explains that the silence generated there can
have major disruptive effects on online communication, from interfering
with team collaboration, to creating misunderstandings.

This can be viewed through the lens of expectation violation theory as
well, as described in \textcite{kalman11} and \textcite{sheldon06}.
Surprisingly, they find different results; both concluded that the
reward valence of the person violating chronemic norms changes how that
violation is perceived, but \textcite{sheldon06} found that low-reward
violaters were more simply more negatively perceived that high-reward
violaters. \textcite{kalman11} found a more complex interaction,
although both studies agreed that the norm violation was perceived
negatively. Part of that is the context, as both the studies were
centered around work. That lines up with Walther's claims that task
focused messages with high latencies were percieved negatively. These
studies validate SIP theory by suggesting that chronemics are a form of
nonverbal communication that participants tend to read into; cues
filtered out theory would expect participants to ignore or discount the
chronemics because the communication is happening in cyberspace.

\hypertarget{chronemics-applications}{%
\section{Chronemics Applications}\label{chronemics-applications}}

Part of the reason cmc chronemics are so interesting is that people
build the systems that these interactions occur in; to me that means
that if we can discover patterns in cmc chronemics that are well
understood, people can apply those patterns when building cmc systems.
For example, if a short response latency automatically meant that two
communicators were closer to each other, that would be useful for
measuring relationship strength and recommending future relational links
in a social network. As far as I can tell, no such simple application
exists. Instead, it seems that chronemics are being analyzed, but there
isn't all that much interest in it as a measurement. Part of this is
because of contamination by other correlated variables, which will be
discussed later. First, there are some promising uses of chronemics,
starting with an apparent correlation with personality type discovered
by \textcite{kalman13}.

This study evaluates a measure called interpost pause. It builds on the
findings of \textcite{kalman11} and SIP theory by using chronemics as
nonverbal cues that participants use in place of normal face to face
cues. \textcite{kalman13} found that people who were more extraverted
``exhibited shorter interpost pauses,'' and that pairs who trusted each
other less had longer interpost pauses (p.~16). Interestingly, the
correlation with trust was stronger than the correlation with
extraversion. \textcite{walther95} had previously observed that a faster
response suggested more equal standing between the two participants,
which matches these results well.

Unfortunately, it's hard to know why these two results exist. For trust,
\textcite{kalman13} suggests it may come down to the assumption that
lying is harder than telling the truth (and therefore should take
longer), but it may also be that people just prefer their conversation
partner to respond quickly and the dislike of slow responses bleeds into
their trust for the person. For extraversion, \textcite{kalman13}
suggests this is mostly because extraverted people talk more and with
less hesitation than intraverted people. Essentially, the same thing
happens in face to face communication because that's just the way people
are. To me, this is incredible because it opens the possibility of
studying those attributes with access to chat logs and without surveys
which can be difficult to sample properly. For example, if Facebook
wants to find out which of its users trust each other, it could
conceivably make a computer model depending on interpost pause in
messaging to find that out. Previously, they would have had to do a
survey, then write a model comparing attributes in the survey to user
attributes. Those attributes may not have yielded any correlation at
all.

The idea that interpost pause or measurable chronemics of any kind could
be very useful is something that for some reason, I couldn't find many
studies on. In particular, I was surprised I didn't see studies using it
to weight relationships and generate tie strength measurements. Facebook
and other social medias generally recommend ``friends'' or other social
links algorithmically, and in order to do that, they usually need
measures of a user's current relationships and their strengths. Link
recommender systems for social media are incredibly important and
profitable, and if you can generate better measures of relationships,
you should get a better recommendation. First, we'll discuss some cases
where chonemics were able to improve tie strength prediction.

\textcite{arnaboldi13} is a good example of this. In particular, they
use five chronemic variables, when most other studies have only one or
two. The most effective were ``number of days since last communication,
the frequency of contact (bidirectional and related to incoming
interactions) and the number of days since first communication''
\autocite[ 1137]{arnaboldi13}. Recency of communication was a great
predictor of tie strength, validating the idea that tie strength should
be predicted by chronemics. Unfortunately, most researchers (like
\textcite{servia-rodriguez14}) only look at recency and duration of
relationship. While this makes sense from a running time standpoint, it
seems that interpost pause or time frame of most communication (like
from \textcite{walther95}) could be useful.

Here we'll discuss studies by \textcite{marsden84} and
\textcite{liberatore17} that may provide an answer to why chronemics
isn't used more. Tie strength measurements in social networks (in
general rather than the cmc specific meaning) have existed for a long
time. One study which provides an example of that is
\textcite{marsden84}. It considers multiple variables for constructing
tie strength in social networks. In particular, they consider
measurements for duration and frequency of contact, in face to face
interaction. The most important part of the study is that they find
issues with using duration and frequency of contact, because there are
so many confounding variables with those two measures. One example they
discuss is neighbors - you may see your neighbor every day, and be
dragged into an hour long conversation with them every day, and still
dislike them or feel indifferent towards them. I think this type of
confounding tends to make researchers apprehensive when they consider
attempting studies like Kalman et al's 2013 study. There are other
explanations though.

One can be found in \textcite{liberatore17}, in which a meta-study is
conducted on the different ways to generate tie strength in cmc social
networks. This is the measurement that I would expect to see chronemics
factor into. The study does consider chronemics with one measure:
Duration (of relationship) \autocite{liberatore17}. Unfortunately, many
of the approaches in \textcite{liberatore17} don't really consider
chronemics outside of that factor. One of the reasons may be the
difficulty of generating the data for every user; the studies which did
include chronemics of individual pair conversations typically used a lot
of variables, and still didn't consider interpost pause. To get an idea
of the problem with scale and runtime, it's worth mentioning that what
Facebook considers its ``fast slice-and-dice data store'' contains 100
terabytes of data \autocite{bronson14}. Unfortunately, the more
variables used to generate tie strength, the worse the model can be to
use, because the model can risk over fitting the data and being quite
slow and expensive to use. In particular, if Facebook wants to generate
tie strengths for their entire network, it will prefer models that are
simpler because they take less computation time and are more reliable.

\hypertarget{conclusion}{%
\section{Conclusion}\label{conclusion}}

Researchers have refuted theories that cmc doesn't have chronemic
nonverbal cues. A cues-filtered-out theory was introduced by
\textcite{sproull91}, and was refuted in \textcite{walther95}.
Researchers have documented their existence and influence on
conversations, and have used them to some extent, although there is
still much to be exploited in terms of their use. SIP theory, as
described in \textcite{kim05}, explains that cmc is in fact personal and
that participants perceive nonverbal cues. Many chronemic phenomena, as
defined in \textcite{walther95} are in fact perceived as nonverbal cues
and influence positive and negative general affection, dominance, and
team collaboration \autocite{walther95,doring09,kalman05,kalman11}.
Researchers have been able to put chronemic cues to work to some extent
to generate tie strengths \autocite{arnaboldi13,servia-rodriguez14}.
However, chronemic cues can be influenced by confounding variables which
can limit their use \autocite{marsden84}. Few tie strength models use
chronemic cues beyond coarse duration of relationship or recency of
contact \autocite{liberatore17} despite chronemics offering a wealth of
information on user personality and dyad trust \autocite{kalman13}. To
look ahead a bit, the research in \textcite{liberatore17} and
\textcite{kalman13} suggests that there is an opportunity to try to
predict tie strength with interpost pause and aggregate time of
messaging.

\printbibliography

\end{document}
